Crunching Large Graphs with Commodity Processors

### 2011-04-19 doc/hotpar2011/softxmt-hotpar2011.tex
insert new cache,delaynet figs (not final versions); draft their captions; rewrite cache pressure and latency experiment parts of eval, but still needs polish.  Section 4.1.2 still needs changes.

Figure~\ref{fig:listwalk-totalconc} shows the result. Each point represents
the maximum rate pointers are traversed for a given number of
concurrent offered references. We see a maximum rate of 277 million references per second (\mrps), which agrees with the measurements found by \cite{Mandal:2010}. This rate is achieved when the number of offered references is 36. Note that a single core cannot support this level of
memory concurrency; the maximum reference rate for a single core is
107M. We believe this limit is due to the core having only enough {\em line fill
  buffers} \cite{nehalem:arch} to store 10 concurrent private cache misses.

@@

% cache pressure figure
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.42\textwidth]{figures/cache-pressure-edited.pdf}
  \end{center}
  \vspace{-0.3in}
  \caption{Cache pressure with coroutines, on six cores.}
  \label{fig:cache-pressure}
\end{figure}

We varied the size of the local working set from 2KB
to 4MB, and ran both 0 and 32 local updates for each pointer
traversal. We measured the overall reference rate, including both the local updates and remote references.

Figure~\ref{fig:cache-pressure} shows the result. As we would expect,
when there are no references to the local structures, the reference rates grow with the number of coroutines just as in
Figure~\ref{fig:multi-green}. For 32 local updates per traversal, while performance
decreases for large working set sizes and large numbers of coroutines,
we see speedup as the number of coroutines increases up to 128KB working set size with 8 coroutines per
core. \todo{preempt questions about drop at 16 which is much less than the about 50 needed per core by saying that the drop is likely the same drop effect seen earlier, which is likely to be prefetch squashing since it happens at number of line fill buffers}

@@

To simulate the performance of pointer chasing in a multi-node system
we access the other chip's memory and also have the scheduler of our coroutine library include a delay
before a coroutine can be rescheduled, imitating the network transit
delay even though we are referencing local memory. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
\cite{infiniband, mellanox:site}, and thus estimate a 3 $\mu$s round-trip
delay.

%simulated delay plot
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.47\textwidth]{figures/multi-green-delay7800-edited.pdf}
  \end{center}
  \vspace{-0.3in}
  \caption{Pointer chasing with simulated network delay. Each core ran up to 167 coroutines, and each coroutine made 1 to 4 concurrent references.}
  \label{fig:network-delay}
\end{figure}

The results of the experiment are shown in
Figure~\ref{fig:network-delay}. We are still able to reach a
near-maximum rate of 273 \mrps\footnote{The throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support. This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, but at least 2
concurrent references per coroutine are required. Running with one
reference per coroutine achieves 236 \mrps.

These results suggest that the runtime should be able to tolerate
large latency well enough to exceed the max issue rates of both modern
network interfaces (about 100 \mrps) \cite{mellanox:press} and the QPI
link used to reach the network interface.

@@

\todo{also have results that remote cores over QPI plus delay 3us can make it up to the 175M, with sufficient number of coroutines}

@@

There has also been much work on user level threading, including the First-Class
User Level Threading project \cite{ult} and Capriccio \cite{capriccio}. We have a
different goal; we want many lightweight contexts to tolerate memory
latency.

>>>>

Figure~\ref{fig:listwalk-totalconc} shows the result. Each point represents
the maximum rate pointers are traversed for a given number of
concurrent offered references. We see a maximum rate of 277 million references per second (\mrps), which agrees with the measurements in \cite{Mandal:2010}. This rate is achieved when the number of offered references is 36. Note that a single core cannot support this level of
memory concurrency; the maximum reference rate for a single core is
107 \mrps. We believe this limit is due to the core having only enough {\em line fill
  buffers} \cite{nehalem:arch} to store 10 concurrent private cache misses.

@@

% cache pressure figure
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{figures/cachepressurenewview-edited.pdf}
  \end{center}
  \vspace{0.0in}
  \caption{Cache pressure with coroutines, on six cores. The vertical axis is remote reference rate. (a) with 8 coroutines per core. (b) with 48 coroutines per core, about the number required for the network rate in simulated delay experiments. }
  \label{fig:cache-pressure}
\end{figure}

We varied the size of the per-thread local working set from 2KB
to 4MB, and ran 0 to 32 local updates for each pointer
traversal. We measured the remote reference rate (i.e. accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).

Figure~\ref{fig:cache-pressure}a shows the result for 8 coroutines per core. 
With 1 to 4 updates per remote reference for smaller working sets, the runtime achieves near maximum throughput. As working set size increases, cache misses in the local space become a bottleneck. When there are several updates to large local spaces, the data of other coroutines is likely to be kicked out and performance suffers.

%As we would expect, when there are no references to the local structures, the reference rates grow with the number of coroutines, just as in Figure~\ref{fig:multi-green}. For 32 local updates per traversal, while performance decreases for large working set sizes and large numbers of coroutines, we see speedup as the number of coroutines increases up to 128KB working set size with 8 coroutines per core.

@@

To simulate the performance of pointer chasing in a multi-node system
we referenced the second chip's memory and also had coroutine scheduler include a delay
before a coroutine can be rescheduled, imitating the network transit
delay even though we are accessing local memory. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
\cite{infiniband, mellanox:site}, and thus estimate a 3 $\mu$s round-trip
delay.

%simulated delay plot
%TODO: replace with the thru QPI version
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.47\textwidth]{figures/delay7800-remote-edited.pdf}
  \end{center}
  \vspace{-0.3in}
  \caption{Pointer chasing across QPI link and with simulated network delay. Used six cores and varied the number of coroutines per core.}
  \label{fig:network-delay}
\end{figure}

The results of the experiment are shown in
Figure~\ref{fig:network-delay}. The runtime can still reach a
maximum rate of 173 \mrps\footnote{The throughput keeps increasing up to around 564 total concurrent references offered (at which point QPI is saturated), far more than the 36 we know the chip to support. This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, saturating the QPI link. It reaches the estimated network rate of 100 \mrps with about 53 coroutines per core.

These results suggest that the runtime should be able to tolerate
large latency well enough to exceed the max issue rates of both modern
network interfaces (about 100 \mrps) \cite{mellanox:press} and the QPI
link used to reach the network interface.
\todo{integrate this paragraph with previous ones instead}

Figure~\ref{fig:cache-pressure}b shows the results of the cache pressure experiment for 48 coroutines per core, about the number of contexts needed to reach the network rate in the delay experiment. We observe that when performing up to 8 local updates per remote reference, smaller working sets still allow the desired throughput. When performing 1 to 2 updates, we can have working sets of up to 32KB per coroutine while maintaining throughput.

@@

@@

There has also been much work on user level threading, including the First-Class
User Level Threading project \cite{ult} and Capriccio \cite{capriccio}. We have a
different goal \todo{we didnt say what their goal is}; we want many lightweight contexts to tolerate memory
latency.
